Using Cell Phone Keyboards is (\NP) Hard

### 2010-01-16 article.tex
Typo fix
Fixed an egregious write-o in accounting for the number of sequences

If we examine a cell phone keyboard (Figure~\ref{keypic}) then we can see that
there are some terrible choices being made.  The frequently-occurring letters
'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
keyboard can do better than this.  If we just reversed the order of the letters
on the 7 key, from `pqrs' to `srqp', the number of button presses required
would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
that every button press takes an equal amount of time, then this corresponds
the users spending $8.7\%$ less time entering their text messages.  In an
effort to find out how low this can go, we define the problem {\sc
BasicCellPhoneTyping} based on the template problem on
page~\pageref{probtemplate}.

@@

\begin{figure}
\begin{algorithm}
\Algorithm{GreedyBasicCellPhoneTyping}~$(A, k, W)$\+\\
    // $A$ is the set of letters \\
    // $k$ is the number of keys \\
    // $W$ is a set of pairs of words and their corresponding frequencies \\
    lettercount $\gets$ new\_map() \\
    \For $c \in A$\+\\
        lettercount$[c] \gets 0$\-\\
    \For (word, frequency) $\in W$\+\\
        \For $c \in$ word \+\\
            lettercount$[c] \gets $ lettercount$[c] +$frequency\-\-\\
    ordered $\gets \left[ ({\rm lettercount}[c], c)~\For c \in {\rm lettercount} \right]$\\
    sort(ordered)\\
    keys $\gets$ a array of length $k$ where each element is an empty list\\
    \For $i \gets 0 \ldots length($ordered$)$\+\\
        (count, char) $\gets$ ordered$[i]$\\
        append(char, keys$[i mod k]$)\-\\
    \Return keys
\end{algorithm}
\caption{The greedy algorithm for Problem~\ref{bcpt}.}
\label{greedyalg}
\end{figure}

To solve {\sc GreedyBasicCellPhoneTyping} we construct a provably optimal
greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
letters, and then assigning letters to keys round-robin style in the order from
most-popular to least-popular.  To prove optimality, we invoke an exchange argument.

@@

When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.  When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in our corpus.  In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.  Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.  The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.  Formally, we extend Problem~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:

@@

Because our problems are \NP-hard, we move on to combinatorial enumeration and experimental results.  The number of partitions of a set of size $n$, which we denote $p(n)$ is actually relatively small.  In particular, $p(26) = 2,436$\footnote{\url{http://www.wolframalpha.com/input/?i=PartitionsP[26]}}, which is easily enumerated and tested.

>>>>

If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we can see that
there are some terrible choices being made.  The frequently-occurring letters
'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
keyboard can do better than this.  If we just reversed the order of the letters
on the 7 key, from `pqrs' to `srqp', the number of button presses required
would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
that every button press takes an equal amount of time, then this corresponds
the users spending $8.7\%$ less time entering their text messages.  In an
effort to find out how low this can go, we define the problem {\sc
BasicCellPhoneTyping} based on the template problem on
page~\pageref{probtemplate}.

@@

\begin{figure}
\begin{algorithm}
\Algorithm{GreedyBasicCellPhoneTyping}~$(A, k, W)$\+\\
    // $A$ is the set of letters \\
    // $k$ is the number of keys \\
    // $W$ is a set of pairs of words and their corresponding frequencies \\
    lettercount $\gets$ new\_map() \\
    \For $c \in A$\+\\
        lettercount$[c] \gets 0$\-\\
    \For (word, frequency) $\in W$\+\\
        \For $c \in$ word \+\\
            lettercount$[c] \gets $ lettercount$[c] +$frequency\-\-\\
    ordered $\gets \left[ ({\rm lettercount}[c], c)~\For c \in {\rm lettercount} \right]$\\
    sort(ordered)\\
    keys $\gets$ a array of length $k$ where each element is an empty list\\
    \For $i \gets 0 \ldots length($ordered$)$\+\\
        (count, char) $\gets$ ordered$[i]$\\
        append(char, keys$[i mod k]$)\-\\
    \Return keys
\end{algorithm}
\caption{The greedy algorithm for Prob.~\ref{bcpt}.}
\label{greedyalg}
\end{figure}

To solve {\sc BasicCellPhoneTyping} we construct a provably optimal
greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
detailed in Fig.~\ref{greedyalg}, and involves creating a histogram of the
letters, and then assigning letters to keys round-robin style in the order from
most-popular to least-popular.  To prove optimality, we invoke an exchange argument.

@@

When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.  When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in our corpus.  In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.  Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.  The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.  Formally, we extend Prob.~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:

@@

Because our problems are \NP-hard, we move on to combinatorial enumeration and experimental results.  The number of partitions of a sequence of size $n$ into $k$ subsequences is equal to $\binom{n-1}{k-1}$, and in the particular case of the 26 letter alphabet and the eight keys available on a cell phone keyboard, we find that there are $\binom{26-1}{8-1} = 480,700$ possible sequences.
